---
title: "Chapter 1"
format: html
---

# Main Text

```{r}
library(ggplot2)
library(janitor)
library(car)
library(xtable)
```

```{r}
dWHO <- read.csv("datasets/WHOLifeExpectancy.csv", header = T)
dim(dWHO)
```

```{r}
dWHO <- clean_names(dWHO)
```


```{r}
slm1 <- lm(formula = life_expectancy ~ year, data = dWHO)
summary(slm1)
```

```{r}
attributes(slm1)
```

```{r}
slm1$coefficients
```

```{r}
#| fig-width: 10
#| fig-height: 5
plot(life_expectancy ~ year, xlab = "Year", ylab = "Life Expectancy", dWHO)
abline(slm1, lwd = 3, col = "red")
```

```{r}
residualPlots(slm1)
```

```{r}
qqPlot(slm1)
```

```{r}
mlm0 <- lm(life_expectancy ~ year * status, data = dWHO)
summary(mlm0)
```

```{r}
anova(mlm0)
```

```{r}
mlm1 <- lm(life_expectancy ~ year + status, data = dWHO)
summary(mlm1)
```

```{r}
knitr::kable(
  round(xtable(mlm1), digits = 4),
  caption = "Summary of the Parameter Estimation",
  booktabs = TRUE
)
```

```{r}
#| fig-width: 6
#| fig-height: 4
plot(
  life_expectancy ~ year,
  xlab = "Year",
  ylab = "Life Expectancy",
  las = 1,
  pch = 3:4,
  col = c("blue", "red"),
  data = dWHO
)
abline(mlm1$coef[1], mlm1$coef[2], lwd = 3, lty = 8, col = "blue")
abline(mlm1$coef[1] + mlm1$coef[3], mlm1$coef[2], lwd = 3, lty = 4, col = "red")
```

```{r}
dt <- dWHO[complete.cases(dWHO), ]
dim(dWHO)
```

```{r}
dim(dt)
```

```{r}
mlm2 <- lm(life_expectancy ~ ., data = dt[, -1])
summary(mlm2)
```

```{r}
mlm3 <- step(object = mlm2, direction = "both")
```

```{r}
summary(mlm3)
```

```{r}
#| fig-width: 5
#| fig-height: 3
hist(
  mlm3$residuals,
  main = "Residual Distribution",
  xlab = "Residuals from mlm3"
)
```

```{r}
#| fig-width: 6
#| fig-height: 4
qqnorm(mlm3$residuals, main = "Normal Q-Q Plot of Residuals")
qqline(mlm3$residuals)
```

```{r}
#| fig-width: 6
#| fig-height: 4
qqPlot(mlm3, distribution = "norm")
```

p-value is small, so we reject the null hypothesis of normality of residuals. This is mostly because of the outliers shown in the Q-Q plot.

```{r}
shapiro.test(mlm3$residuals)
```

```{r}
#| fig-width: 10
#| fig-height: 6
car::residualPlots(mlm3, layout = c(3, 5))
```

# Exercises

## Exercise 1.1

- status
  - gdp: development status may amplify/dampen the effect of wealth
- status
  - schooling: education's impact likely differs by Developed vs. Developing

```{r}
mlm_interaction <- lm(
  life_expectancy ~ . + status:gdp + status:schooling,
  data = dt[, -1]
)
summary(mlm_interaction)
```

```{r}
summary(mlm3)
```

```{r}
anova(mlm3, mlm_interaction)
```

The interaction model has a significantly better fit, suggesting that the effects of gdp and schooling on life expectancy differ by development status.

## Exercise 1.2

```{r}
#| fig-width: 5
#| fig-height: 3
hist(
  mlm_interaction$residuals,
  main = "Residual Distribution",
  xlab = "Residuals from mlm_interaction"
)
```

```{r}
#| fig-width: 6
#| fig-height: 4
qqnorm(mlm_interaction$residuals, main = "Normal Q-Q Plot of Residuals")
qqline(mlm_interaction$residuals)
```

```{r}
#| fig-width: 6
#| fig-height: 4
qqPlot(mlm_interaction, distribution = "norm")
```

p-value is small, so we reject the null hypothesis of normality of residuals. This is mostly because of the outliers shown in the Q-Q plot.

```{r}
shapiro.test(mlm_interaction$residuals)
```

## Exercise 1.3                                                                                              
`residualPlots(mlm3)` showed non-linear patterns for several variable, suggesting that quadratic terms may improve fit. We add them to `mlm3` and compare.                                          

```{r}                                                                                                       
mlm3_quad <- lm(
  life_expectancy ~ year +
    status +
    adult_mortality +
    infant_deaths +
    alcohol +
    percentage_expenditure +
    bmi +
    under_five_deaths +
    total_expenditure +
    diphtheria +
    hiv_aids +
    thinness_5_9_years +
    income_composition_of_resources +
    schooling +
    I(adult_mortality^2) +
    I(alcohol^2) +
    I(percentage_expenditure^2) +
    I(under_five_deaths^2) +
    I(diphtheria^2) +
    I(hiv_aids^2) +
    I(thinness_5_9_years^2) +
    I(income_composition_of_resources^2),
  data = dt[, -1]
)
summary(mlm3_quad)
```

```{r}                                                                                                       
#| fig-width: 10                                                                                             
#| fig-height: 6                                                                                             
car::residualPlots(mlm3_quad, layout = c(3, 5))
```                                                                                                          

```{r}                                                                                                       
AIC(mlm3, mlm3_quad)
```

## Chapter 1 Notes

### Shapiro-Wilk Test — W Statistic

- W ranges from 0 (completely non-normal) to 1 (perfectly normal)
- p < 0.05 means statistically non-normal, but the W value indicates the *practical* departure
- With large n (e.g., n > 1000), Shapiro-Wilk becomes very sensitive — even trivial deviations yield p < 0.05
- W = 0.99 with large n is practically normal; the Central Limit Theorem protects inference

| Sample Size | Sensitivity |
|---|---|
| Small (n=30) | W ≈ 0.93 or lower needed for p < 0.05 |
| Large (n=1,649) | Even W ≈ 0.999 can yield p < 0.05 |

### Normality Assumption in Practice

- Normality is the *least critical* regression assumption
- For inference (p-values, CIs): the CLT covers you at large n
- For prediction intervals: normality matters more, especially in the tails

| Assumption | Why It Matters |
|---|---|
| **Linearity** | Biased estimates if violated |
| **Homoscedasticity** | Unreliable standard errors if violated |
| **Independence** | Can completely invalidate inference if violated |
| **Normality of residuals** | Least critical, especially at large n |

### AIC Interpretation

- Lower AIC = better model (think golf score)
- Penalises model complexity (more parameters = higher AIC penalty)
- Large drops (e.g., >10 units) indicate meaningful improvement

### `residualPlots` — What to Focus On

- Focus on the *original variable rows*, not `I(x^2)` rows
- Tests whether non-linearity in each predictor is resolved by added terms
- `I(x^2)` rows flagging significance means even higher-order terms or transformations are needed

### Polynomials vs. Transformations

- `x^2` only captures symmetric (parabolic) non-linearity
- For skewed, exponential, or plateau relationships, transformations are more appropriate:

| Relationship Type | Recommended Transformation |
|---|---|
| Right-skewed | `log(x)` or `log(x+1)` if zeros are present |
| Plateau effect | `sqrt(x)` |
| Exponential decay | `log(x)` |

### Revealed Non-Linearity After Model Improvement

- Adding quadratic terms reduces residual noise (lower Residual Standard Error, RSE — the average size of model residuals)
- It's the standard deviation of the residuals, calculated as:

  $$RSE = \sqrt{\frac{\sum(y_i - \hat{y}_i)^2}{n - p - 1}}$$

  - $y_i - \hat{y}_i$ = residuals
  - $n$ = number of observations
  - $p$ = number of predictors (the $-1$ corrects for degrees of freedom)
- Smaller residuals → higher statistical power → previously hidden non-linearities become detectable
- Variables that appeared linear before may show significance after model refinement
- This is expected and not a mistake — it reflects improved model precision

### Iterative Modelling Workflow

- residualPlots → add quadratic terms or transformations → re-run step() to drop redundant variables → residualPlots again → repeat until convergence
- Convergence = no significant patterns in residualPlots, or improvements become negligible
- Perfect normality/linearity is rarely achievable with real-world data — document remaining issues as limitations